---
title: "Reproducible Report on COVID19 Data"
author: "Dan Wesely"
output:
  html_document: default
  pdf_document: default
date: "2025-02-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(tidyverse)
library(lubridate)
```

## Introduction

COVID 19 data project, based heavily on examples provided in class.


## Loading Data

Load global and US data.

```{r load}
global_deaths <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/refs/heads/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv")
us_cases <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/refs/heads/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv")
global_cases <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/refs/heads/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")
us_deaths <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/refs/heads/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv")

#global_deaths <- read_csv("time_series_covid19_deaths_global.csv")
#us_cases <- read_csv("time_series_covid19_confirmed_US.csv")
#global_cases <- read_csv("time_series_covid19_confirmed_global.csv")
#us_deaths <- read_csv("time_series_covid19_deaths_US.csv")

uid <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/refs/heads/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv") %>% select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))
#uid <- read_csv("UID_ISO_FIPS_LookUp_Table.csv") %>% select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

```

## Tidying Data

Create tables of global and US cases and deaths.

```{r pivotglobal}
global_cases_pivot <- global_cases %>% pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long), names_to = "date", values_to = "cases") %>% select(-c(Lat,Long))
global_deaths_pivot <- global_deaths %>% pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long), names_to = "date", values_to = "deaths") %>% select(-c(Lat,Long))
```

```{r pivotus}
us_cases_pivot <- us_cases %>% pivot_longer(cols = -(UID:Combined_Key), names_to = "date", values_to = "cases") %>% select(Admin2:cases) %>% mutate(date = mdy(date)) %>% select(-c(Lat, Long_))
us_deaths_pivot <- us_deaths %>% pivot_longer(cols = -(UID:Population), names_to = "date", values_to = "deaths") %>% select(Admin2:deaths) %>% mutate(date = mdy(date)) %>% select(-c(Lat, Long_))
us <- us_cases_pivot %>% full_join(us_deaths_pivot)
```

## Merge Tables

Combine global deaths and cases counts with population of the area.

```{r combine_pop}
global <- global_cases_pivot %>% full_join(global_deaths_pivot) %>% rename(Country_Region = `Country/Region`, Province_State = `Province/State`) %>% mutate(date = mdy(date))
global <- global %>% unite("Combined_Key", c(Province_State, Country_Region), sep = ", ", na.rm = TRUE, remove = FALSE)
global <- global %>% left_join(uid, by = c("Province_State", "Country_Region")) %>% select(-c(UID, FIPS)) %>% select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)
```

Combine US deaths and cases counts with population of the area.

```{r bystate}
us_by_state <- us %>% group_by(Province_State, Country_Region, date) %>% summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>% mutate(deaths_per_mill = deaths *1000000 / Population) %>% select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population) %>% ungroup()

us_by_state <- us_by_state %>% mutate(new_cases = cases - lag(cases), new_deaths = deaths - lag(deaths))

us_totals <- us_by_state %>% group_by(Country_Region, date) %>% summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>% mutate(deaths_per_mill = deaths *1000000 / Population) %>% select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>% ungroup()

us_totals <- us_totals %>% mutate(new_cases = cases - lag(cases), new_deaths = deaths - lag(deaths))
```

## Visualization

Plot cases and deaths over time, in US and in New York. Cumulative counts increase over time.


```{r viz}
us_totals %>% filter(cases > 0) %>% ggplot(aes(x = date)) + geom_line(aes(y = cases, color = "cases")) + geom_point(aes(y = cases, color = "cases")) + scale_y_log10() + theme(legend.position="bottom", axis.text.x = element_text(angle = 90)) + labs(title = "COVID19 in US: Cases and Deaths", y = NULL)  + geom_line(aes(color = "deaths", y = deaths)) + geom_point(aes(x = date, y = deaths, color = "deaths"))

state <- "New York"
  us_by_state %>% filter(Province_State == state) %>% filter(cases > 0) %>% ggplot(aes(x = date, y = cases)) + geom_line(aes(color = "cases", y = cases)) + geom_point(aes(color = "cases", y = cases)) + geom_line(aes(y = deaths, color = "deaths")) + geom_point(aes(y = deaths, color = "deaths")) + scale_y_log10() + theme(legend.position="bottom", axis.text.x = element_text(angle = 90)) + labs(title = str_c("COVID19 in ", state), y = NULL)

```

## Delta

Plot new cases and new deaths, rather than cumulative total.

```{r newcasesviz}

us_totals %>% ggplot(aes(x = date, y = new_cases)) + geom_line(aes(color = "new_cases")) + geom_point(aes(color = "new_cases")) + geom_line(aes(y = new_deaths, color = "new_deaths")) + geom_point(aes(y = new_deaths, color = "new_deaths")) + scale_y_log10() + theme(legend.position="bottom", axis.text.x = element_text(angle = 90)) + labs(title = "covid19 in US: New Cases", y = NULL)
```

## Rates

Convert counts to rates, using available population data.

```{r perthou}
us_state_totals <- us_by_state %>% group_by(Province_State) %>% summarize(deaths = max(deaths), cases = max(cases), population = max(Population), cases_per_thou = 1000* cases / population, deaths_per_thou = 1000 * deaths / population) %>% filter(cases >0, population > 0)

us_state_totals %>% slice_min(deaths_per_thou, n = 10)

us_state_totals %>% slice_max(deaths_per_thou, n = 10)
```

## Model

Fit a linear model to the data to predict the number of deaths per thousand people, based on the number of cases per thousand people.

```{r model}
mod <- lm(deaths_per_thou ~ cases_per_thou, data = us_state_totals)
summary(mod)
us_state_totals %>% slice_min(deaths_per_thou)
us_state_totals %>% slice_max(deaths_per_thou)
us_state_totals %>% mutate(pred = predict(mod))
us_tot_w_pred <- us_state_totals %>% mutate(pred = predict(mod))
us_tot_w_pred %>% ggplot() + geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") + geom_point(aes(x = cases_per_thou, y = pred), color = "red")
```

## Bias

Differences in reporting processes and infrastructure in the different states could have caused case and death rates to be skewed. Different state resources may have resulted in more or less accurate identification of COVID as well, which could again affect rates.


```{r}
sessionInfo()
```

